A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-10-23 imaging_demo/ms.tex
cleaning up some math notation

Interferometers measure {\it visibilities} -- the degree of coherence -- between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging. This expression can be further simplified with the sky brightness, 
$I(\hat{\mathbf{s}}) \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})= \left\langle 
E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t $, and defining the 
antenna pair sky power response function (or the primary beam), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the primary 
beam, and uncorrelated noise terms which we group into $\widetilde{n}_p$,
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

@@

Following the matrix
notation of \citet{mor11}, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{B}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,I(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $I(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{B}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{E}^\star(\mathbf{a})\widetilde{E}(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

Using the optimal map-making formalism \citep{teg97a,teg97b}, a software
holography image is formed using \citep{mor09}:
\begin{align}
  I^\prime(\hat{\mathbf{s}}) &= \mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{u})\,\widetilde{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v},\mathbf{v})\,\mathbf{m}(\mathbf{v}) \label{eqn:dirty-image-FX}
\end{align}
where the measured visibilities are weighted by the inverse of the system noise,
followed by a gridding process using the holographic antenna power response as 
the gridding kernel, followed by a Fourier transform to create an image
$I^\prime(\hat{\mathbf{s}})$. This is the optimal estimate of the true image 
$I(\hat{\mathbf{s}})$ given the visibility measurements.

The intermediate step of gridding with the antenna power response can be 
expressed as a convolution of a data vector generated by gridding the electric fields 
directly with the antenna illumination pattern.
\begin{align}
&\widetilde{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v})=\nonumber\\
&\qquad\left\langle \left[\widetilde{\mathbf{W}}^\textrm{T}_a(\mathbf{r},\mathbf{a})\, \widetilde{\mathbf{N}}^{-1}\!(\mathbf{a},\mathbf{a})\, \widetilde{\mathbf{E}}(\mathbf{a})\right]
\ast\left[\widetilde{\mathbf{W}}_a(\mathbf{r},\mathbf{a})\, \mathbf{N}^{-1}\!(\mathbf{a},\mathbf{a})\, \mathbf{E}^\star(\mathbf{a})\right]\right\rangle_t\label{eqn:e-field-conv}
\end{align}

We can then use the multiplication-convolution theorem to move the convolution in 
Equation \ref{eqn:e-field-conv} to a square after the Fourier transform in Equation 
\ref{eqn:dirty-image-FX}.
\begin{align}
  I^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\widetilde{\mathbf{W}}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}^{-1}(\mathbf{a},\mathbf{a})\,\widetilde{E}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as
that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna
electric fields are weighted by the antenna noise, weighted and gridded by the
antenna aperture kernel, Fourier transformed and finally squared to obtain the
same image estimated that would have been obtained using equation~
\ref{eqn:dirty-image-FX}.

@@

Figure~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\widetilde{E}_a$ in matrix
notation) for each of the antennas. Each of the complex antenna gains are
calibrated to correct the corresponding electric field spectra. These calibrated
electric fields are gridded using an antenna-based gridding convolution function
after which it is spatially Fourier transformed and squared to obtain images
for every time stamp. These images are then time-averaged to obtain the
accumulated image $I(f)$ ($I(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas which are then Fourier transformed by the F-engine to produce
    electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and
    gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series
    are imaged to produce an images $I_1(f)\ldots I_\textrm{N}(f)$. These images
    are time-averaged to obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

@@

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{e})$ is described
as a transformation matrix from the space of measured electric fields
($\mathbf{e}$) to the antenna array grid denoted by $\mathbf{r}$. Since each
antenna occupies a footprint typically the size of its aperture,
$\mathbf{M}(\mathbf{r},\mathbf{e})$, which is generally of size
$N_\textrm{grid}\times N_\textrm{a}$, reduces to a sparse block-diagonal matrix
with only $N_\textrm{a}$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. Figure~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

The antenna array aperture illumination over the entire grid,
$\widetilde{W}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{W}(\mathbf{r}) &= \sum_a \widetilde{W}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{e})\,\mathcal{I}(\mathbf{e}),
\end{align}
where, $\mathcal{I}(\mathbf{e})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{W}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra. Thus,
\begin{align}
  \widetilde{E}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{e})\,E(\mathbf{e}),
\end{align}
where $E(\mathbf{e})$ denotes the spectra of measured antenna electric fields.
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.

>>>>

Interferometers measure {\it visibilities} -- the degree of coherence between
electric fields measured by a pair of antennas \citep{van34,zer38,tho01}.
A visibility, $\widetilde{V}_p$, can be written as:
\begin{align}
  \widetilde{V}_p = &\left\langle \widetilde{E}_a\widetilde{E}_b^\star \right\rangle_t \label{eqn:cc-vis}\hfill\\
  		= &\left\langle \left[ \int {W}_a(\hat{\mathbf{s}})\,E(\hat{\mathbf{s}})\,e^{-i2\pi\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}}\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_a \right] \right. \nonumber\\
		&\qquad \times \left.\left[ \int {W}^\star_b(\hat{\mathbf{s'}})\,E^\star (\hat{\mathbf{s'}})\,e^{i2\pi\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}}\,\dif^2\hat{\mathbf{s'}} + \widetilde{n}^\star_b \right] \right\rangle_t \\
                  =& \iint {W}_a(\hat{\mathbf{s}}) {W}^\star_b(\hat{\mathbf{s'}}) \left\langle E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t e^{-i2\pi(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})}\,\dif^2\hat{\mathbf{s}}\,\dif^2\hat{\mathbf{s'}},
\end{align}
where we have brought the time average into the integral under the assumption that 
the aperture illumination pattern does not change over the time-scale of the 
averaging. This expression can be further simplified with the sky brightness, 
$I(\hat{\mathbf{s}}) \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})= \left\langle 
E(\hat{\mathbf{s}}) E^\star(\hat{\mathbf{s'}}) \right\rangle_t $, and defining the 
antenna pair sky power response function (or the primary beam), 
$B_p(\hat{\mathbf{s}})~\equiv~{W}_a(\hat{\mathbf{s}})~{W}^\star_b(\hat{\mathbf{s}})$. 
The result is the visibility expressed in terms of the sky brightness, the primary 
beam, and uncorrelated noise terms which we group into $\widetilde{n}_p$,
\begin{equation}
\widetilde{V}_p = \int e^{-i2\pi\mathbf{u}_p\!\cdot\,\hat{\mathbf{s}}}\,B_p(\hat{\mathbf{s}})\,I(\hat{\mathbf{s}})\,\dif^2\hat{\mathbf{s}} + \widetilde{n}_p,
\end{equation}
where the baseline coordinate $\mathbf{u}_p=\mathbf{r}_a-\mathbf{r}_b$ is the 
vector separation between the two antennas. This signifies that the visibility 
($\widetilde{V}_p$) measured between a
pair of antennas ($p$) is obtained by the multiplying the sky brightness
$I(\hat{\mathbf{s}})$ by the antenna power response $B(\hat{\mathbf{s}})$ and
Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to $uv$
coordinates, which are then sampled at the locations of the antenna spacings (or
baselines), namely, $\mathbf{u}_p$, and added to the receiver noise $n_p$.

@@

From here we adopt the matrix notation of \citet{mor11}, where vectors are 
represented with single coordinates, and matrices are represented by two 
coordinates denoting the spaces the operator transforms between. 
In this notation, the above measurement equation can be expressed as:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})\,\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})\,\mathbf{I}(\hat{\mathbf{s}}) + \mathbf{n}(\mathbf{v}),
\end{align}
where the sky brightness $\mathbf{I}(\hat{\mathbf{s}})$ is Fourier transformed using
$\mathbf{F}(\mathbf{u},\hat{\mathbf{s}})$ and the resultant spatial coherence
function is weighted and summed using the antenna power response,
$\widetilde{\mathbf{B}}(\mathbf{v},\mathbf{u})$ in $uv$-space sampled at the baseline
location to obtain the measured visibilities:
\begin{align}
  \mathbf{m}(\mathbf{v}) &= \left\langle \widetilde{\mathbf{E}}^\star(\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}^\prime)\right\rangle_t, \label{eqn:matrix-cc-vis}
\end{align}
where $\mathbf{m}(\mathbf{v})$ denotes visibilities measured by cross-correlating
measured antenna electric fields over all possible pairs of $\mathbf{a}$ and
$\mathbf{a}^\prime$. It is the same as equation~\ref{eqn:cc-vis} written in
matrix notation.

Using the optimal map-making formalism \citep{teg97a,teg97b}, a software
holography image is formed using \citep{mor09}:
\begin{align}
  \mathbf{I}^\prime(\hat{\mathbf{s}}) &= \mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{u})\,\widetilde{\mathbf{B}}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v},\mathbf{v})\,\mathbf{m}(\mathbf{v}) \label{eqn:dirty-image-FX}
\end{align}
where the measured visibilities are weighted by the inverse of the system noise,
followed by a gridding process using the holographic antenna power response as 
the gridding kernel, followed by a Fourier transform to create an image
$\mathbf{I}^\prime(\hat{\mathbf{s}})$. This is the optimal estimate of the true image 
$\mathbf{I}(\hat{\mathbf{s}})$ given the visibility measurements.

The intermediate step of gridding with the antenna power response can be 
expressed as a convolution of a data vector generated by gridding the electric fields 
directly with the antenna illumination pattern.
\begin{align}
&\widetilde{\mathbf{B}}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v} ,\mathbf{v})\,\mathbf{m}(\mathbf{v})=\nonumber\\
&\qquad\left\langle \left[\widetilde{\mathbf{W}}^\textrm{T}_a(\mathbf{r},\mathbf{a})\, \widetilde{\mathbf{N}}^{-1}\!(\mathbf{a},\mathbf{a})\, \widetilde{\mathbf{E}}(\mathbf{a})\right]
\ast\left[\widetilde{\mathbf{W}}_a(\mathbf{r},\mathbf{a})\, \mathbf{N}^{-1}\!(\mathbf{a},\mathbf{a})\, \mathbf{E}^\star(\mathbf{a})\right]\right\rangle_t\label{eqn:e-field-conv}
\end{align}

We can then use the multiplication-convolution theorem to move the convolution in 
Equation \ref{eqn:e-field-conv} to a square after the Fourier transform in Equation 
\ref{eqn:dirty-image-FX}.
\begin{align}
  \mathbf{I}^\prime(\hat{\mathbf{s}}) &= \left\langle \left|\,\mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{r})\,\widetilde{\mathbf{W}}^\textrm{T}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{N}}^{-1}(\mathbf{a},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a})\,\right|^2\right\rangle_t. \label{eqn:dirty-image-MOFF}
\end{align}
The term inside the angular brackets before squaring has a very similar form as
that in equation~\ref{eqn:dirty-image-FX}. It signifies that the measured antenna
electric fields are weighted by the antenna noise, weighted and gridded by the
antenna aperture kernel, Fourier transformed and finally squared to obtain the
same image estimated that would have been obtained using equation~
\ref{eqn:dirty-image-FX}.

@@

Figure~\ref{fig:MOFF-flowchart} shows the flowchart for MOFF imaging. The
propagated electric fields are shown on the left at different time stamps,
$t_1\ldots t_\textrm{M}$. At each time stamp, the electric fields measured by
antennas are denoted by $E_1(t)\ldots E_\textrm{N}(t)$. The F-engine performs a
temporal Fourier transform on the electric field time-series to obtain electric
field spectra $E_1(f)\ldots E_\textrm{N}(f)$ ($\widetilde{\mathbf{E}}(\mathbf{a})$ in matrix
notation) for each of the antennas. Each of the complex antenna gains are
calibrated to correct the corresponding electric field spectra. These calibrated
electric fields are gridded using an antenna-based gridding convolution function
after which it is spatially Fourier transformed and squared to obtain images
for every time stamp. These images are then time-averaged to obtain the
accumulated image $I(f)$ ($\mathbf{I}(\hat{\mathbf{s}})$ in matrix notation).
\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_flowchart.eps}
  \caption{A flowchart of MOFF imaging in EPIC. The propagated electric fields
    shown on the left are measured as time-series $E_1(t)\ldots E_\textrm{N}(t)$
    by the antennas which are then Fourier transformed by the F-engine to produce
    electric field spectra $E_1(f)\ldots E_\textrm{N}(f)$. They are calibrated and
    gridded. The gridded electric fields $E_\textrm{g}(f)$ from each time series
    are imaged to produce an images $I_1(f)\ldots I_\textrm{N}(f)$. These images
    are time-averaged to obtain the final image $I(f)$.}
  \label{fig:MOFF-flowchart}
\end{figure}

@@

In the most general case, this gridding kernel could contain information on the
$w$-projection effect, and even other time-dependent ionospheric effects. For a
stationary antenna array in the absence of any time-dependent effects, this
mapping must only be determined once in the antenna array coordinate frame. The
antenna-to-grid mapping matrix, $\mathbf{M}(\mathbf{r},\mathbf{a})$ is described
as a transformation matrix from the space of measured electric fields by the antennas
($\mathbf{a}$) to the antenna array grid denoted by the coordinate $\mathbf{r}$. Since each
antenna occupies a footprint typically the size of its aperture,
$\mathbf{M}(\mathbf{r},\mathbf{a})$, which is generally of size
$N_\textrm{grid}\times N_\textrm{a}$, reduces to a sparse block-diagonal matrix
with only $N_\textrm{a}$ blocks and roughly $N_\textrm{k}$ non-zero entries per
block. Figure~\ref{fig:a2g-mapping} illustrates the antenna-to-grid mapping
matrix and the grid containing the mapped aperture footprints of the antennas.

@@

The antenna array aperture illumination over the entire grid,
$\widetilde{\mathbf{W}}(\mathbf{r})$, is obtained by a projection of the individual
antenna aperture illuminations:
\begin{align}\label{eqn:gridding-convolution}
  \widetilde{\mathbf{W}}(\mathbf{r}) &= \sum_a \widetilde{\mathbf{W}}_a(\mathbf{r}-\mathbf{r}_a) \\
                            &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\mathcal{I}(\mathbf{a}),
\end{align}
where, $\mathcal{I}(\mathbf{a})$ is a row of ones. This is achieved using
efficient multiplication with the sparse matrix created in the antenna-to-grid
mapping process. Unless $\widetilde{\mathbf{W}}(\mathbf{r})$ includes time-dependent
effects of the ionosphere or the instrument, it needs to be computed just once
for the entire observation. However, the gridding of electric fields must be
computed at every readout of the electric field spectra.
\begin{align}
  \widetilde{\mathbf{E}}(\mathbf{r}) &= \mathbf{M}(\mathbf{r},\mathbf{a})\,\widetilde{\mathbf{E}}(\mathbf{a}),
\end{align}
% The sentence below was already stated earlier.
%In practice the grid is set to be a power of 2 such that the grid spacing is at
%most $\lambda/2$ even at the highest frequency in the observing band.
